Quantitative Auditing of AI Fairness with Differentially Private Synthetic Data

### 2024-12-05 paper/paper.tex
title
abstract wip

%%
%% This is file `sample-manuscript.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `all,proceedings,bibtex,manuscript')
%%
%% IMPORTANT NOTICE:
%%
%% For the copyright see the source file.
%%
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-manuscript.tex.
%%
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%%
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[manuscript,screen,review,anonymous]{acmart}
% \documentclass[acmsmall,sigconf]{acmart}

%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

@@

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% For managing citations, it is recommended to use bibliography
%% files in BibTeX format.
%%
%% You can then either use BibTeX with the ACM-Reference-Format style,
%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
%% support for advanced citation of software artefact from the
%% biblatex-software package, also separately available on CTAN.
%%
%% Look at the sample-*-biblatex.tex files for templates showcasing
%% the biblatex styles.
%%

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}

@@

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Quantitative Auditing of Fairness Measures with Differentially Private Synthetic Data}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  abstract
\end{abstract}

@@

\ccsdesc[500]{General and reference~Metrics}
\ccsdesc[500]{Security and privacy~Usability in security and privacy}
\ccsdesc[300]{Information systems~Data mining}
\ccsdesc[300]{Computing methodologies~Machine learning}
\ccsdesc[300]{Computing methodologies~Artificial intelligence}
\ccsdesc[300]{Security and privacy~Privacy-preserving protocols}
\ccsdesc[300]{Information systems~Information systems applications}
\ccsdesc[300]{Mathematics of computing~Contingency table analysis}
% %%
% \begin{CCSXML}
% <ccs2012>
%  <concept>
%   <concept_id>00000000.0000000.0000000</concept_id>
%   <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
%   <concept_significance>500</concept_significance>
%  </concept>
%  <concept>
%   <concept_id>00000000.00000000.00000000</concept_id>
%   <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
%   <concept_significance>300</concept_significance>
%  </concept>
%  <concept>
%   <concept_id>00000000.00000000.00000000</concept_id>
%   <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
%   <concept_significance>100</concept_significance>
%  </concept>
%  <concept>
%   <concept_id>00000000.00000000.00000000</concept_id>
%   <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
%   <concept_significance>100</concept_significance>
%  </concept>
% </ccs2012>
% \end{CCSXML}

% \ccsdesc[500]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
% \ccsdesc[300]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
% \ccsdesc{Do Not Use This Code~Generate the Correct Terms for Your Paper}
% \ccsdesc[100]{Do Not Use This Code~Generate the Correct Terms for Your Paper}

% %%
% %% Keywords. The author(s) should pick words that accurately describe
% %% the work being presented. Separate the keywords with commas.
% \keywords{Do, Not, Us, This, Code, Put, the, Correct, Terms, for,
%   Your, Paper}

@@

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

@@

We looked at several publicly available datasets, such as adult\cite{adult_2,Kaggle_Adult_Census_Income}, COMPAS\cite{larson2016propublica,Kaggle_COMPAS_Dataset}, and diabetes\cite{diabetes_34,Kaggle_Diabetes_Prediction}. The adult dataset comes from the 1994 census in the United States and contains about 30000 individuals. The COMPAS dataset comes from an investigative report by ProPublica of the COMPAS criminal recidivism assessment system and contains about 7000 individuals. The diabetes dataset comes from the hospital readmission data published in the 1994 AI in Medicine journal and contains about 100000 individuals. Since these datasets all have binary outcomes, we did not consider the case of non-binary outcomes in the experiments.

@@

\begin{figure}
  \centering
  \begin{subfigure}[b]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{adult_mst}
      \caption{Markov random field for marginals of the adult dataset.}
      \label{fig:adult_mst}
      \Description{Markov random field for marginals of the adult dataset.}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{compas_mst}
      \caption{Markov random field for marginals of the COMPAS dataset.}
      \label{fig:compas_mst}
      \Description{Markov random field for marginals of the COMPAS dataset.}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{diabetes_mst}
      \caption{Markov random field for marginals of the diabetes dataset.}
      \label{fig:diabetes_mst}
      \Description{Markov random field for marginals of the COMPAS dataset.}
  \end{subfigure}
      \caption{Markov random field for marginals of the experimented datasets.}
      \label{fig:msts}
\end{figure}

@@

For the adult income dataset, the shape of the maximum spanning tree is very shallow, almost resembling a star; it has one internal node and all but one of the leaves have a depth of one. After introducing edges according to our heuristic, we observed an increase in the pairwise edges of the leaves, forming many 3-cliques and two 4-cliques. The resulting graph is shown in Figure~\ref{fig:adult_mst}.

@@

\begin{table}[h]
\caption{
    Fairness measures experiment results of the adult dataset.
    With heuristic, average difference is $0.0787$, and fit time is 924m15s.
    Without heuristic, average difference is $0.0775$, and fit time is 17m1s.
}
\label{tab:adult_score}
\begin{tabular}{ccccccc}
\toprule
\textbf{Measure} & \textbf{Original} & \textbf{Synthetic} & \textbf{Difference} & \textbf{Synthetic(No Heuristic)} & \textbf{Difference(No Heuristic)} \\
\midrule
Demographic Parity  & 0.1933 & 0.1079 & 0.0853 & 0.0114 & 0.1818 \\
Accuracy Eqaulity   & 0.0250 & 0.1399 & 0.1149 & 0.0169 & 0.0080 \\
Equalized Odds 1    & 0.0175 & 0.0870 & 0.0695 & 0.0060 & 0.0115 \\
Equalized Odds 2    & 0.0114 & 0.0622 & 0.0508 & 0.0303 & 0.0189 \\
Accuracy Equality 1 & 0.1941 & 0.1466 & 0.0474 & 0.0315 & 0.1626 \\
Accuracy Equality 2 & 0.1102 & 0.1450 & 0.0348 & 0.0299 & 0.0803 \\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{table}[h]
\caption{
    Fairness measures experiment results of the diabetes dataset.
    With heuristic, average difference is $0.0067$, and fit time is 1m44s.
    Without heuristic, average difference is $0.0078$, and fit time is 1m24s.
}
\label{tab:diabetes_score}
\begin{tabular}{cccccc}
\toprule
\textbf{Measure} & \textbf{Original} & \textbf{Synthetic} & \textbf{Difference} & \textbf{Synthetic(No Heuristic)} & \textbf{Difference(No Heuristic)} \\
\midrule
Demographic Parity  & 0.0135 & 0.0038 & 0.0096 & 0.0015 & 0.0119 \\
Accuracy Eqaulity   & 0.0077 & 0.0011 & 0.0065 & 0.0011 & 0.0065 \\
Equalized Odds 1    & 0.0000 & 0.0006 & 0.0006 & 0.0010 & 0.0010 \\
Equalized Odds 2    & 0.0086 & 0.0106 & 0.0020 & 0.0094 & 0.0008 \\
Accuracy Equality 1 & 0.0133 & 0.0090 & 0.0042 & 0.0059 & 0.0074 \\
Accuracy Equality 2 & 0.0216 & 0.0041 & 0.0175 & 0.0019 & 0.0197 \\
\bottomrule
\end{tabular}
\end{table}

>>>>

\documentclass[manuscript,screen,review,anonymous]{acmart}
% \documentclass[acmsmall,sigconf]{acmart}

@@

@@

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Quantitative Auditing of AI Fairness with Differentially Private Synthetic Data}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Ensuring fairness in AI systems is crucial, but auditing these systems often requires sensitive data, which raises security and privacy concerns. To address this, we introduce a framework that utilizes Differentially Private synthetic data for fairness evaluations. This approach allows auditors to assess AI systems without exposing sensitive information. Through experiments on datasets like Adult, COMPAS, and Diabetes, we compare fairness metrics of synthetic and real data. Our results ...
\end{abstract}

@@

\ccsdesc[500]{General and reference~Metrics}
\ccsdesc[500]{Security and privacy~Usability in security and privacy}
\ccsdesc[300]{Information systems~Data mining}
\ccsdesc[300]{Computing methodologies~Machine learning}
\ccsdesc[300]{Computing methodologies~Artificial intelligence}
\ccsdesc[300]{Security and privacy~Privacy-preserving protocols}
\ccsdesc[300]{Information systems~Information systems applications}
\ccsdesc[300]{Mathematics of computing~Contingency table analysis}

\keywords{AI, fairness, auditing, differential privacy, synthetic data}

@@

\maketitle

@@

We looked at several publicly available datasets, such as Adult\cite{adult_2,Kaggle_Adult_Census_Income}, COMPAS\cite{larson2016propublica,Kaggle_COMPAS_Dataset}, and Diabetes\cite{diabetes_34,Kaggle_Diabetes_Prediction}. The Adult dataset comes from the 1994 census in the United States and contains about 30000 individuals. The COMPAS dataset comes from an investigative report by ProPublica of the COMPAS criminal recidivism assessment system and contains about 7000 individuals. The Diabetes dataset comes from the hospital readmission data published in the 1994 AI in Medicine journal and contains about 100000 individuals. Since these datasets all have binary outcomes, we did not consider the case of non-binary outcomes in the experiments.

@@

\begin{figure}
  \centering
  \begin{subfigure}[b]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{adult_mst}
      \caption{Markov random field for marginals of the Adult dataset.}
      \label{fig:adult_mst}
      \Description{Markov random field for marginals of the Adult dataset.}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{compas_mst}
      \caption{Markov random field for marginals of the COMPAS dataset.}
      \label{fig:compas_mst}
      \Description{Markov random field for marginals of the COMPAS dataset.}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{diabetes_mst}
      \caption{Markov random field for marginals of the Diabetes dataset.}
      \label{fig:diabetes_mst}
      \Description{Markov random field for marginals of the COMPAS dataset.}
  \end{subfigure}
      \caption{Markov random field for marginals of the experimented datasets.}
      \label{fig:msts}
\end{figure}

@@

For the Adult income dataset, the shape of the maximum spanning tree is very shallow, almost resembling a star; it has one internal node and all but one of the leaves have a depth of one. After introducing edges according to our heuristic, we observed an increase in the pairwise edges of the leaves, forming many 3-cliques and two 4-cliques. The resulting graph is shown in Figure~\ref{fig:adult_mst}.

@@

\begin{table}[h]
\caption{
    Fairness measures experiment results of the Adult dataset.
    With heuristic, average difference is $0.0787$, and fit time is 924m15s.
    Without heuristic, average difference is $0.0775$, and fit time is 17m1s.
}
\label{tab:adult_score}
\begin{tabular}{ccccccc}
\toprule
\textbf{Measure} & \textbf{Original} & \textbf{Synthetic} & \textbf{Difference} & \textbf{Synthetic(No Heuristic)} & \textbf{Difference(No Heuristic)} \\
\midrule
Demographic Parity  & 0.1933 & 0.1079 & 0.0853 & 0.0114 & 0.1818 \\
Accuracy Eqaulity   & 0.0250 & 0.1399 & 0.1149 & 0.0169 & 0.0080 \\
Equalized Odds 1    & 0.0175 & 0.0870 & 0.0695 & 0.0060 & 0.0115 \\
Equalized Odds 2    & 0.0114 & 0.0622 & 0.0508 & 0.0303 & 0.0189 \\
Accuracy Equality 1 & 0.1941 & 0.1466 & 0.0474 & 0.0315 & 0.1626 \\
Accuracy Equality 2 & 0.1102 & 0.1450 & 0.0348 & 0.0299 & 0.0803 \\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{table}[h]
\caption{
    Fairness measures experiment results of the Diabetes dataset.
    With heuristic, average difference is $0.0067$, and fit time is 1m44s.
    Without heuristic, average difference is $0.0078$, and fit time is 1m24s.
}
\label{tab:diabetes_score}
\begin{tabular}{cccccc}
\toprule
\textbf{Measure} & \textbf{Original} & \textbf{Synthetic} & \textbf{Difference} & \textbf{Synthetic(No Heuristic)} & \textbf{Difference(No Heuristic)} \\
\midrule
Demographic Parity  & 0.0135 & 0.0038 & 0.0096 & 0.0015 & 0.0119 \\
Accuracy Eqaulity   & 0.0077 & 0.0011 & 0.0065 & 0.0011 & 0.0065 \\
Equalized Odds 1    & 0.0000 & 0.0006 & 0.0006 & 0.0010 & 0.0010 \\
Equalized Odds 2    & 0.0086 & 0.0106 & 0.0020 & 0.0094 & 0.0008 \\
Accuracy Equality 1 & 0.0133 & 0.0090 & 0.0042 & 0.0059 & 0.0074 \\
Accuracy Equality 2 & 0.0216 & 0.0041 & 0.0175 & 0.0019 & 0.0197 \\
\bottomrule
\end{tabular}
\end{table}
